Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-16 paper-ieee-trans.tex
added ack

The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project. Conclusions are drawn in Section \ref{sec:con}.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would had been accomplished through the
development and maintenance of some operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
the production infrastructures. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defined a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, will provide guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of scope, objective fulfilment,
documentation completeness, etc. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also 
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} -- as the online source code repository --,
Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the {\sl INDIGO--DataCloud}???s application repository.

@@

After more that a decade of experiences collected throughout the reviewed
EU-funded software development and e-Infrastructure management projects, it
becomes apparent that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. The gradual adoption of software engineering insights, such as
DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
collaborative tools, extending their capabilities to bring along tight
integrations amongst other services, matched the requirements of such practices
and procedures, facilitating in a great manner their implementation in real scenarios.

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre-production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process --, alleviates the likelihood of discovering uncovered
issues at production.

@@

{\sl INDIGO--DataCloud} project has received funding from the European Research
Council (ERC) under the European Union's Horizon 2020 research and innovation
programme (grant agreement n?? 777536).

>>>>

The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section \ref{sec:con}.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
Table ~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This has been accomplished through the
development and maintenance of some operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
the production infrastructures. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers, such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defined a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, provides guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} promotes automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation completeness. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also 
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github}, as the online source code repository,
Jenkins \cite{jenkins}, the event-response CI system, and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the {\sl INDIGO--DataCloud} application repository.

@@

After more than a decade of experiences collected throughout the reviewed
EU-funded software development and e-Infrastructure management projects, it
becomes apparent that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. The gradual adoption of software engineering insights, such as
DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
collaborative tools, extending their capabilities to bring along tight
integrations amongst other services, matched the requirements of such practices
and procedures, facilitating in a great manner their implementation in real scenarios.

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre-production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process, alleviates the likelihood of discovering uncovered
issues at production.

@@

{\sl DataGrid - Research and Technological Development for an International Data Grid} project 
has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
FP5-IST - Programme for research, technological development and demonstration on a "User-friendly information society, 1998-2002" (IST--2000--25182E)

{\sl EGEE - Enabling grids For E-science} project
has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
 FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)

{\sl EGEE - Enabling grids For E-science--II} project
 has received funding from the European Commission - Community Research and Development Information System (CORDIS) under 
 FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high-capacity and high-speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)

{\sl EGEE - Enabling grids For E-science--III} project
  has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e-Science Grid infrastructures", (INFSO--RI--222667)

{\sl INDIGO--DataCloud} project has received funding from the European Research
Council (ERC) under the European Union's Horizon 2020 research and innovation
programme (GA--777536).
